---
title: "Statistical Inference: ToothGrowth"
author: "EHarris"
date: "6/28/2017"
output: html_document
---

## Packages for Course 6 Project: Tooth Growth
```{r setup, include = FALSE, echo = FALSE, results = "hide"}
ipak <- function(pkg){
      new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
      if (length(new.pkg)) 
            install.packages(new.pkg, dependencies = TRUE)
      sapply(pkg, require, character.only = TRUE)
}

### Package Usage
packages <- c("plyr", "dplyr", "data.table", "dtplyr", "lubridate", "ggplot2", "scales",
                  "reshape2", "knitr", "R.cache", "stringr", "gtools", "quantreg")
ipak(packages)
```

## Overview  
Using the R dataset ToothGrowth, we are asked to evaluate, using confidence intervals and/or hypothesis testing, tooth growth by supp and dose.  The first step is to determine whether there has been growth, then whether the growth was signifiantly different from others. An important assumption is that growth cannot be negative.  The negative growth would be an indication of other conditions or trauma such that we would remove from the analysis.  


## Load Data  
Given that this is an R dataset we can simply use data(ToothGrowth) to load the dataset into the current enviromnent.  
```{r load_dataset, include = TRUE, echo = TRUE}
data(ToothGrowth)
```

## Exploratory Data Analysis  
Let's take a closer look at the datsset to understand the contents of the dataset.  Two key parameters for the analysis are supp and dose.  We want to these are not NULL for an observation.  If we observe such a NULL value, we will need to determine whether we can make a reasonable substitution or the observation is removed from analysis.

### Overall summary view of dataset
```{r dataset_overview, include = TRUE, echo = TRUE, results = "asis"}
str(ToothGrowth)
summary(ToothGrowth)
table(ToothGrowth$supp, ToothGrowth$dose)
```

### Unique values for 'supp' and 'dose'
```{r unique_values, include = TRUE, echo = TRUE, results = "hold"}
unique(ToothGrowth$supp)
unique(ToothGrowth$dose)
```

### Boxplot: Length of tooth for different supplements & dosing levels
With a limited number of variables, the boxplot provides a good vehicle to illustrate the data.  Not only only can we  compare tooth lenght by supplement (supp) and dose, we capture insight to the range of values (length) for each combination of supplement and dose.
```{r plot1_scatterplot, fig.keep = "high", fig.show = "asis", fig.path = 'figure/'}
ggplot(ToothGrowth, aes(x = supp, y = len)) +
      geom_boxplot(aes(fill = supp)) +
      facet_wrap(~dose)
```

Looking at information from the exploratory data analysis, there are two key observations: 1) higher doses produce greater tooth growth; and 2) 'OJ' produces tooth growth equal to or better than 'VC' at all observed dose levels, particularly at lower doses. As there are 10 observations for each supp-dose combination, we will use T-statistics to evaluate whether 'OJ' results are significantly better th 'VC' and whether dose levels produce materially different growth.


## Confidence Intervals & Hypothosis Testing
As previously stated, there are two variables of interest, supplement('VC', 'OJ') and dose(0.5, 1.0, and 2.0).  Our initial testing will be to look at the supplement.  There may be two questions to be answered here.  First, does 'OJ' promote greater tooth growth than 'VC'?  This would be across all dosing levels.  Once we answer this question, we may want to evaluate the supplement within each dosing level.

### T-test: Difference in Tooth Growth by Supplement
The first step is to split data by supplement('supp').  We will use the t.test() function in R to evaluate whether the tooth length is longer when the supplement 'OJ' is used compared to 'VC' across all dose levels.  The boxplot might indicate that the supplement 'OJ' promotes greater tooth growth than 'VC'.  We will use a one-sides (right) to evaluate
```{r supp_ttl_difference, include = TRUE, echo = TRUE, results = "hold"}
vc_ttl <- select(subset(ToothGrowth, supp == "VC"), supp, len)
oj_ttl <- select(subset(ToothGrowth, supp == "OJ"), supp, len)

t.test(oj_ttl$len,vc_ttl$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

The result of this test is that t = 1.9153 falls within the 95% confidence interval of [0.4682687, Inf] and a p-value = 0.03032. These statistics indicate that we reject a NULL hypothesis that there is no significant difference in favor of the alternative hypothesis, $H_a$, that 'OJ' promotes greater tooth growth than 'VC', at a 95% confidence interval.

### T-test: Difference in Tooth Growth by Supplement with a dose level
We are focus on the difference in tooth growth produced by a supplement and dose level.  Before applying t.test(), we will subset data to reflect ths specified populations.  We will create 6 data subsets, each represent a specific supp and dose.  For example, vc05 represents data for the supp 'VC' at a 0.5 dose level.  Similarly, oj10 reprsents data for the supp 'OJ' and dose level of 1.0.

##### T-test: Greater Tooth Growth with the supplement 'OJ' versus 'VC' at dose = 0.5
```{r supp_dose05_difference, include = TRUE, echo = TRUE, results = "hold"}
vc05 <- select(subset(ToothGrowth, supp == "VC" & dose == 0.5), supp, len)
vc10 <- select(subset(ToothGrowth, supp == "VC" & dose == 1.0), supp, len)
vc20 <- select(subset(ToothGrowth, supp == "VC" & dose == 2.0), supp, len)
oj05 <- select(subset(ToothGrowth, supp == "OJ" & dose == 0.5), supp, len)
oj10 <- select(subset(ToothGrowth, supp == "OJ" & dose == 1.0), supp, len)
oj20 <- select(subset(ToothGrowth, supp == "OJ" & dose == 2.0), supp, len)

t.test(oj05$len,vc05$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

We are testing whether we would reject the NULL hypothesis that tooth growth for 'VC' and 'OJ' are similar at a dose = 0.5.  The alternative hypothesis, $H_a$, is that 'OJ' promotes greater tooth growth at dose = 0.5.  Performing a t-test produced the following statistics: t = 3.1697 falls between the 95% confidence interval [2.34604, Inf] and a p-value = 0.003179. Both statistics suggest that we reject the NULL hypothesis in favor of the alternative hypothesis, supp 'OJ' promotes greater tooth growth than 'VC'.  We recommend that the NULL hypothesis is rejected.

##### T-test: Greater Tooth Growth with the supplement 'OJ' versus 'VC' at dose = 1.0
```{r supp_dose10_difference, include = TRUE, echo = TRUE, results = "hold"}
t.test(oj10$len,vc10$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

As with dose = 0.5, we are testing the NULL hypothesis that tooth growth for the supplements 'OJ' and 'VC' are not significantly different for a dose = 1.0.  The alternative hypothesis, $H_a$, is that 'OJ' promotes greater tooth growth than 'VC'.  The t-test statistics include: t = 4.0328 with confidence interval [3.356258, Inf] and p-value = 0.0005192.  Both statistics suggest that the NULL hypothesis be rejected in favor of the alternative hypothesis that the supplement 'OJ' promotes greater tooth growth than 'VC'.  We recommend that the NULL hypothesis be rejected.


##### T-test: Difference in Tooth Growth by Supplement with dose = 2.0
```{r supp_dose20_difference, include = TRUE, echo = TRUE, results = "hold"}
t.test(oj20$len,vc20$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

Based on the boxplot we would not expect that there is any significant difference in tooth growth at the dose = 2.0 level.  The test statistics confirm this belief.  With t = -0.046136, it falls within the 95% confidence interval [-3.1335, Inf].  The p-value = 0.5181.  Using the t-value alone, you might recommend rejecting the NULL hypothesis that tooth growth for 'OJ' and "VC' using dose = 2.0 is not significantly different.  The p-value, however, suggest that we accept the NULL hypothesis.  Based on the results of t.test, we recommend that the NULL hypothesis be accepted.

### T-test: Difference in Tooth Growth due to dose
The overall goal is to determine whether increasing the dose of a supplement('supp') results in greater tooth growth.  Because results are not different at the 95% confidence interval we might be able to compare data without regard to the type of supplement.  We have elected to look at the effects of dose within each supplement. For each supplement, separately, we will compare tooth length for the different dose levels.  The initial comparison is tooth length with a dose = 2.0 to a dose = 0.5.

##### Tooth Growth: Supp ('OJ') by comparing dose [0.5, 2.0]
```{r tooth_growth_dose_max_oj, include = TRUE, echo = TRUE, results = "hold"}
t.test(oj20$len,oj05$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

A couple statistics from the t.test.  A t = 7.817 with a confidence interval [9.94845, Inf] and p-value = 6.619e-07 (0.0000006619)
For supp = 'OJ', this test does suggest that we reject the NULL hypothesis that there is no difference in tooth growth between the dose levels, 0.5 & 2.0.  The t value = -7.817 falls outside the 95% confidence interval range [-16.335421, -9.324759]. Additionally, we have a p-value = 1.324e-06 (0.000001324).  Both statistics indicate a rejection of teh NULL hypothesis


##### Tooth Growth: Supp ('VC') by comparing dose [0.5, 2.0]
```{r tooth_growth_dose_max_vc, include = TRUE, echo = TRUE, results = "hold"}
t.test(vc20$len,vc05$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

Results for the supplement = 'VC' are consistent with 'OJ'.   t-value = 10.388 clearly lies outside the 95% confidence interval [-21.90151, -14.41849]. This, along with p-value = 4.682e-08 (0.00000004682), suggest that we reject the NULL hypothesis.  Again, this confirms that, at a 95% confidence level, there is a difference between tooth growth at a dose = 0.5 and 2.0.

#### Tooth Growth:  Significant at dose levels?
Although it appears to be clear that there is greater tooth growth between the lower dose = 0.5 and the highest dose = 2.0, we want to compare growth for the interim dose levels, between 0.5 - 1.0; 1.0 - 2.0.

##### Tooth Growth: Supp ('OJ') by comparing interim dose
```{r tooth_growth_dose_interim1_oj, include = TRUE, echo = TRUE, results = "hold"}
t.test(oj10$len,oj05$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

```{r tooth_growth_dose_interim2_oj, include = TRUE, echo = TRUE, results = "hold"}
t.test(oj20$len,oj10$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```


##### Tooth Growth: Supp ('VC') by comparing interim dose
```{r tooth_growth_dose_interim1_vc, include = TRUE, echo = TRUE, results = "hold"}
t.test(vc10$len,vc05$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

```{r tooth_growth_dose_interim2_vc, include = TRUE, echo = TRUE, results = "hold"}
t.test(vc20$len,vc10$len, paired = FALSE, var.equal = FALSE, alternative = "greater")
```
