---
title: "Statistical Inference: ToothGrowth"
author: "EHarris"
date: "6/28/2017"
output: html_document
---

## Packages for Course 6 Project: Tooth Growth
```{r setup, include = TRUE, echo = FALSE, results = "hide"}
ipak <- function(pkg){
      new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
      if (length(new.pkg)) 
            install.packages(new.pkg, dependencies = TRUE)
      sapply(pkg, require, character.only = TRUE)
}

### Package Usage
packages <- c("plyr", "dplyr", "data.table", "dtplyr", "lubridate", "ggplot2", "scales",
                  "reshape2", "knitr", "R.cache", "stringr", "gtools", "quantreg")
ipak(packages)
```

## Overview  
Using the R dataset ToothGrowth, we are asked to evaluate, using confidence intervals and/or hypothesis testing, tooth growth by supp and dose.  The first step is to determine whether there has been growth, then whether the growth was signifiantly different from others. An important assumption is that growth cannot be negative.  The negative growth would be an indication of other conditions or trauma such that we would remove from the analysis.  


## Load Data  
Given that this is an R dataset we can simply use data(ToothGrowth) to load the dataset into the current enviromnent.  
```{r load_dataset, include = TRUE, echo = TRUE}
data(ToothGrowth)
```


## Exploratory Data Analysis  
Let's take a closer look at the datsset to understand the contents of the dataset.  Two key parameters for the analysis are supp and dose.  We want to these are not NULL for an observation.  If we observe such a NULL value, we will need to determine whether we can make a reasonable substitution or the observation is removed from analysis.

### Overall summary view of dataset
```{r dataset_overview, include = TRUE, echo = TRUE, results = "asis"}
str(ToothGrowth)
summary(ToothGrowth)
table(ToothGrowth$supp, ToothGrowth$does)
```

### Unique values for 'supp' and 'dose'
```{r unique_values, include = TRUE, echo = TRUE, results = "asis"}
unique(ToothGrowth$supp)
unique(ToothGrowth$dose)
```

### Plot 1: Boxplot
With a limited number of variables, the boxplot provides a good vehicle to illustrate the data.  Not only only can we  compare tooth lenght by supplement (supp) and dose, we capture insight to the range of values (length) for each combination of supplement and dose.
```{r plot1_scatterplot, fig.keep = "high", fig.show = "asis", fig.path = 'figure/'}
ggplot(ToothGrowth, aes(x = supp, y = len)) +
      geom_boxplot(aes(fill = supp)) +
      facet_wrap(~dose)
```

Looking at information from the exploratory data analysis, there are two key observations: 1) higher doses produce greater tooth growth; and 2) 'OJ' produces tooth growth equal to or better than 'VC' at all observed dose levels, particularly at lower doses. As there are 10 observations for each supp-dose combination, we will use T-statistics to evaluate whether 'OJ' results are significantly better th 'VC' and whether dose levels produce materially different growth.






